---
title: "ggmap section"
subtitle: "Workshop on ggplot"
author: "`r paste0('<b>Lokesh Mano</b> • ',format(Sys.time(), '%d-%b-%Y'))`"
output:
  bookdown::html_document2:
          toc: true
          toc_float: true
          toc_depth: 4
          number_sections: true
          theme: flatly
          highlight: tango
          df_print: default
          code_folding: "none"
          self_contained: false
          keep_md: false
          encoding: 'UTF-8'
          css: "assets/lab.css"
---

```{r, include=FALSE}
hooks = knitr::knit_hooks$get()
hook_foldable = function(type) {
  force(type)
  function(x, options) {
    res = hooks[[type]](x, options)
    
    if (isFALSE(options[[paste0("fold.", type)]])) return(res)
    
    paste0(
      "<details><summary>", type, "</summary>\n\n",
      res,
      "\n\n</details>"
    )
  }
}
knitr::knit_hooks$set(
  output = hook_foldable("output"),
  plot = hook_foldable("plot")
)
```

```{r,child="assets/header-lab.Rmd"}
```

# ggmap

`ggmap` is an R package that makes it easy to retrieve raster map tiles from popular online mapping services like Google Maps and Stamen Maps and plot them using the `ggplot2` framework:

You can find more information about the R package [here](https://github.com/dkahle/ggmap)

Here we will try to see some of the important functions that this package provides and how we can use it display the findings in a nicer way.

To do this exercise, we will use a separate dataset and it can be downloaded [here](data/ggmap_data/ggmap_data.zip). 

If you would like to keep the data tree same as the exercises here, you can extract the zip file inside the data directory. So, the file tree would like this:

* <span style="color:blue">ggplot_geneco_course</span>
  + <span style="color:blue">data</span>
    + archaea_nexus.txt
    + arch_newick.txt
    + counts_deseq2.txt
    + counts_filtered.txt
    + counts_raw.txt
    + counts_vst.txt
    + human_biomaRt_annotation.csv
    + metadata_raw.csv
    + Time_t24_vs_t0.txt
    + Time_t2_vs_t0.txt
    + Time_t6_vs_t0.txt
    + tree_env.tsv
    + tree_hmap.tsv
    + <span style="color:blue">ggmap_data</span>
      + bc_sites.rds
      + bike-ride.csv
      + sisquoc-points.txt


```{r, warning=FALSE, message=FALSE}
if (!requireNamespace("ggmap", quietly = TRUE)){
  install.packages("ggmap", dependencies = TRUE)
}

library(dplyr)
library(ggplot2)
library(reshape2)
library(tidyverse)
library(ggmap)
```