---
title: "ggmap section"
subtitle: "Workshop on ggplot"
author: "`r paste0('<b>Lokesh Mano</b> • ',format(Sys.time(), '%d-%b-%Y'))`"
output:
  bookdown::html_document2:
          toc: true
          toc_float: true
          toc_depth: 4
          number_sections: true
          theme: flatly
          highlight: tango
          df_print: default
          code_folding: "none"
          self_contained: false
          keep_md: false
          encoding: 'UTF-8'
          css: "assets/lab.css"
---

```{r, include=FALSE}
hooks = knitr::knit_hooks$get()
hook_foldable = function(type) {
  force(type)
  function(x, options) {
    res = hooks[[type]](x, options)
    
    if (isFALSE(options[[paste0("fold.", type)]])) return(res)
    
    paste0(
      "<details><summary>", type, "</summary>\n\n",
      res,
      "\n\n</details>"
    )
  }
}
knitr::knit_hooks$set(
  output = hook_foldable("output"),
  plot = hook_foldable("plot")
)
```

```{r,child="assets/header-lab.Rmd"}
```

# Making maps with R

To do this exercise, we will use a separate dataset and it can be downloaded [here](data/ggmap_data/ggmap_data.zip). 

If you would like to keep the data tree same as the exercises here, you can extract the zip file inside the data directory. So, the file tree would like this:

* <span style="color:blue">ggplot_geneco_course</span>
  + <span style="color:blue">data</span>
    + archaea_nexus.txt
    + arch_newick.txt
    + counts_deseq2.txt
    + counts_filtered.txt
    + counts_raw.txt
    + counts_vst.txt
    + human_biomaRt_annotation.csv
    + metadata_raw.csv
    + Time_t24_vs_t0.txt
    + Time_t2_vs_t0.txt
    + Time_t6_vs_t0.txt
    + tree_env.tsv
    + tree_hmap.tsv
    + <span style="color:blue">ggmap_data</span>
      + bc_sites.rds
      + bike-ride.csv
      + sisquoc-points.txt
      + world_population.tsv


Before, we go into `ggmap` package and it's options. Let us take a quick look about some of the other options available for making maps in R.

## maps and mapdata

Here we will see, how some of the traditional R packages `maps` and `mapdata` were used and how we could implement the information from these packages into `ggplot`. Some pointers about these packages are:

* The maps package contains a lot of outlines of continents, countries, states, and counties that have been with R for a long time.
* The mapdata package contains a few more, higher-resolution outlines.
* The maps package comes with a plotting function, but, we will opt to use ggplot2 to plot the maps in the maps package.
* Recall that ggplot2 operates on data frames. Therefore we need some way to translate the maps data into a data frame format the ggplot can use.

Let us start with how to install these packages and look at their simple functions:

```{r, message=FALSE, warning=FALSE}
if (!requireNamespace("maps", quietly = TRUE)){
  install.packages("maps", dependencies = TRUE)
}

if (!requireNamespace("mapdata", quietly = TRUE)){
  install.packages("mapdata", dependencies = TRUE)
}

library(maps)
library(mapdata)
library(ggplot2)
library(dplyr)
library(reshape2)
library(tidyverse)
world <- map_data("world")
head(world)
```

The structure of this data-set could be explained as follows:

* `long` is longitude. Things to the west of the prime meridian are negative.
* `lat` is latitude.
* `order`: This just shows in which order ggplot should “connect the dots”
* `region` and `subregion` tell what region or subregion a set of points surrounds.
* `group`: This is very important! ggplot2’s functions can take a group argument which controls (amongst other things) whether adjacent points should be connected by lines. If they are in the same group, then they get connected, but if they are in different groups then they don’t. 
  + Essentially, having to points in different groups means that ggplot “lifts the pen” when going between them.

To plot the map, in a pretty way:

```{r, message=FALSE, warning=FALSE}
gg1 <- ggplot(data = world, aes(x=long, y = lat)) + 
  geom_polygon(aes(group = group)) + 
  theme(legend.position = "none")
gg1

```

<i class="fas fa-exclamation-circle"></i> Note &nbsp; New Zealand was not missed in this map ;)

<div class="instruction">

* Maps in this format can be plotted with the polygon geom. i.e. using geom_polygon().
* `geom_polygon()` drawn lines between points and “closes them up” (i.e. draws a line from the last point back to the first point)
* You have to map the group aesthetic to the group column
* Of course, x = long and y = lat are the other aesthetics.

You can test it out how the data looks like without using the group aesthetic

```{r, message=FALSE, warning=FALSE, eval=FALSE}
gg1 <- ggplot(data = world, aes(x=long, y = lat)) + 
  geom_polygon() + 
  theme(legend.position = "none")
gg1

```

</div>

From the dataset we have, we can also draw the border lines representing the different countries based on the variable `region`. 

```{r, message=FALSE, warning=FALSE}
gg2 <- ggplot(data = world, aes(x=long, y = lat)) + 
  geom_polygon(aes(group = group, fill = region)) + 
  theme(legend.position = "none")
gg2
```

### Adding points to map

Now, in this huge map, let us try to add some points like our favourite city Lund and few others.

```{r, message=FALSE, warning=FALSE}
labs <- data.frame(
  long = c(13.1629767,11.7537556,16.3100205,80.0689272),
  lat = c(55.7068448,57.7010982,48.220778,13.0478223),
  names = c("LUND", "GBG", "Vienna", "Chennai"),
  stringsAsFactors = FALSE
  )  

gg2 + 
  geom_point(data = labs, aes(x = long, y = lat), color = "black", size = 2) 

```

<i class="fas fa-exclamation-circle"></i> Note &nbsp; You can also import the points from a file and store in a data frame.

### Zooming in

Unlike in normal ggplots, to zoom into a section of the map, if we try to use the `xlim()` and `ylim()` functions:

```{r, message=FALSE, warning=FALSE}
gg2 + 
  geom_point(data = labs, aes(x = long, y = lat), color = "black", size = 2) +
  xlim(c(-30,40)) +
  ylim(c(50,85))
```

You can see that this subsection of the map is not very good, you can notice it especially in the area of Germany and Poland. This is because the latitude and longtitude does not really work like axes on a plot and the structure of the data-set we have here doesn't support it that way. Well, at-least I hope you are not a flat-earther and will try to disprove me here ;)

We will go through a function in `ggmap` to get one of the nicer ways to get a subsection in those measures. But, with the data we have here, it is possible get a subset of countries:

```{r, message=FALSE, warning=FALSE}
scandic <- subset(world, region %in% c("Sweden", "Denmark", "Norway", "Finland", "Iceland", "Faroe Islands"))

gg_scan <- ggplot(data = scandic, aes(x=long, y=lat)) + 
  geom_polygon(aes(group = group, fill = region)) 
gg_scan
```

### Continous data

Let's see how we can add a continous data like the World population statistics on to the above map. We can do this by importing the file with the population information and join it to the data set, we have with `left_join()`

```{r, message=FALSE, warning=FALSE}
pop_info <- read.table("data/ggmap_data/world_population.tsv", header = T, row.names = 1, sep = "\t")
head(pop_info)
```

<i class="fas fa-exclamation-circle"></i> Note &nbsp; Here the values are in the unit of thousands.

```{r, message=FALSE, warning=FALSE}
world_pop <- left_join(world, pop_info, by = "region")

gg_pop <- ggplot(data = world_pop, aes(x=long, y = lat, group = group, fill =PopTotal))
gg_pop + geom_polygon(color = "white") +
  scale_fill_gradientn(colours = rainbow(8))
```


# ggmap

`ggmap` is an R package that makes it easy to retrieve raster map tiles from popular online mapping services like Google Maps and Stamen Maps and plot them using the `ggplot2` framework:

You can find more information about the R package [here](https://github.com/dkahle/ggmap)

Here we will try to see some of the important functions that this package provides and how we can use it display the findings in a nicer way.

```{r, warning=FALSE, message=FALSE}
if (!requireNamespace("ggmap", quietly = TRUE)){
  install.packages("ggmap", dependencies = TRUE)
}

library(ggmap)
```