---
title: "Base and Grid graphics"
subtitle: "Workshop on ggplot"
author: "`r paste0('<b>Lokesh Mano</b> • ',format(Sys.time(), '%d-%b-%Y'))`"
output:
  bookdown::html_document2:
          toc: true
          toc_float: true
          toc_depth: 4
          number_sections: true
          theme: flatly
          highlight: tango
          df_print: default
          code_folding: "none"
          self_contained: false
          keep_md: false
          encoding: 'UTF-8'
          css: "assets/lab.css"
---

```{r,child="assets/header-lab.Rmd"}
```

```{r,include=FALSE}
# data handling
library(dplyr)
#library(tidyr)
#library(stringr)

# plotting
library(ggplot2)

#library(biomaRt) # annotation
#library(DESeq2) # rna-seq
#library(edgeR) # rna-seq
```


# Graphics

## Base

R is an excellent tool for creating graphs and plots. The graphic capabilities and functions provided by the base R installation is called the base R graphics. Numerous packages exist to extend the functionality of base graphics.

We can try out plotting a few of the common plot types. Let's start with a scatterplot. First we create a `data.frame` as this is the most commonly used data object.

```{r}
dfr <- data.frame(a=sample(1:100,10),b=sample(1:100,10))
```

Now we have a dataframe with two continuous variables that can be plotted against each other.

```{r}
plot(dfr$a,dfr$b)
```

This is probably the simplest and most basic plots. We can modify the x and y axis labels.

```{r}
plot(dfr$a,dfr$b,xlab="Variable a",ylab="Variable b")
```

We can change the point to a line.

```{r}
plot(dfr$a,dfr$b,xlab="Variable a",ylab="Variable b",type="b")
```

Let's add a categorical column to our dataframe.

```{r}
dfr$cat <- rep(c("C1","C2"),each=5)
```

And then colour the points by category.

```{r}
# subset data
dfr_c1 <- subset(dfr,dfr$cat == "C1")
dfr_c2 <- subset(dfr,dfr$cat == "C2")

plot(dfr_c1$a,dfr_c1$b,xlab="Variable a",ylab="Variable b",col="red",pch=1)
points(dfr_c2$a,dfr_c2$b,col="blue",pch=2)

legend(x="topright",legend=c("C1","C2"),
       col=c("red","blue"),pch=c(1,2))
```

Let's create a barplot.

```{r}
ldr <- data.frame(a=letters[1:10],b=sample(1:50,10))
barplot(ldr$b,names.arg=ldr$a)
```

## Grid

Grid graphics have a completely different underlying framework compared to base graphics. Generally, base graphics and grid graphics cannot be plotted together. The most popular grid-graphics based plotting library is **ggplot2**.

Let's create the same plot are before using **ggplot2**. Make sure you have the package installed.

```{r}
library(ggplot2)

ggplot(dfr,aes(x=a,y=b,colour=cat))+
  geom_point()+
  labs(x="Variable a",y="Variable b")
```

It is generally easier and more consistent to create plots using the ggplot2 package compared to the base graphics.

Let's create a barplot as well.

```{r}
ggplot(ldr,aes(x=a,y=b))+
  geom_bar(stat="identity")
```


## Images

Let's take a look at saving plots.

## Base graphics

The general idea for saving plots is open a graphics device, create the plot and then close the device. We will use **png** here. Check out `?png` for the arguments and other devices.

```{r,eval=FALSE}
dfr <- data.frame(a=sample(1:100,10),b=sample(1:100,10))

png(filename="plot-base.png")
plot(dfr$a,dfr$b)
dev.off()
```

## ggplot2

The same idea can be applied to ggplot2, but in a slightly different way. First save the file to a variable, and then export the plot.

```{r,eval=FALSE}
p <- ggplot(dfr,aes(a,b)) + geom_point()

png(filename="plot-ggplot-1.png")
print(p)
dev.off()
```

**ggplot2** also has another easier helper function to export images.

```{r,eval=FALSE}
ggsave(filename="plot-ggplot-2.png",plot=p)
```

# Session info

```{r,echo=FALSE}
sessionInfo()
```

__End of document__
