---
title: "Intro to R"
subtitle: "Workshop on ggplot"
author: "`r paste0('<b>Lokesh Mano</b> • ',format(Sys.time(), '%d-%b-%Y'))`"
output:
  bookdown::html_document2:
          toc: true
          toc_float: true
          toc_depth: 4
          number_sections: true
          theme: flatly
          highlight: tango
          df_print: default
          code_folding: "none"
          self_contained: false
          keep_md: false
          encoding: 'UTF-8'
          css: "assets/lab.css"
---

```{r,child="assets/header-lab.Rmd"}
```

```{r,include=FALSE}
# data handling
library(dplyr)
#library(tidyr)
#library(stringr)

# plotting
library(ggplot2)

#library(biomaRt) # annotation
#library(DESeq2) # rna-seq
#library(edgeR) # rna-seq
```

R is a programming language for statistical computing, and data wrangling. It is open-source, widely used in data science, has a wide range of functions and algorithms for graphing and data analyses.

We will use R in this course for generating plots from different biological data that are of higher quality and standard for publications. To do this, you will brush-up your memory on some important aspects of R that are important for this course below:

# Input/Output

Input and output of data and images is an important aspect with data analysis.

## Text

Data can come in a variety of formats which needs to be read into R and converted to an R data type.

Text files are the most commonly used input. Text files can be read in using the function `read.table`. We have a sample file to use: **iris.txt**.

```{r,eval=FALSE}
dfr <- read.table("iris.txt",header=TRUE,stringsAsFactors=F)
```

This reads in a tab-delimited text file with a header. The argument `sep='\t'` is set by default to specify that the delimiter is a tab. `stringsAsFactors=F` setting ensures that character columns are not automatically converted to factors.

It's always a good idea to check the data after import.

```{r,eval=FALSE}
head(dfr)
```

```{r,eval=FALSE}
str(dfr)
```

Check `?read.table` for other wrapper functions to read in text files.

Let's filter this data.frame and create a new dataset.

```{r,eval=FALSE}
dfr1 <- dfr[dfr$Species == "setosa",]
```

And we can write this as a text file.

```{r,eval=FALSE}
write.table(dfr1,"iris-setosa.txt",sep="\t",row.names=F,quote=F)
```

`sep="\t"` sets the delimiter to tab. `row.names=F` denotes that rownames should not be written. `quote=F` specifies that doubles must not be placed around strings.

# Data-Frames

You have probably learnt about `data.frame` in your previous course and this is the most important data structure for generating plots using `ggplot`. So, below you have some quick exercises on how to work with Data-Frames.

Vectors positions can be accessed using `[]`. R follows 1-based indexing, meaning that the indexing starts at 1.

Dataframe or matrix positions can be accessed using `[]` specifying row and column like `[row,column]`.

```{r}
dfr <- data.frame(x = 1:3, y = c("a", "b", "c"))
dfr
dfr[1,]
dfr[,1]
dfr[2,2]
```


The function `cbind()` is used to join two dataframes column-wise.

```{r}
dfr1 <- data.frame(x = 1:3, y = c("a", "b", "c"))
dfr2 <- data.frame(p = 4:6, q = c("d", "e", "f"))
dfr1
dfr2

cbind(dfr1,dfr2)
```

Similarily, `rbind()` is used to join two dataframes row-wise.

```{r}
dfr1 <- data.frame(x = 1:3, y = c("a", "b", "c"))
dfr2 <- data.frame(x = 4:6, y = c("d", "e", "f"))
dfr1
dfr2

rbind(dfr1,dfr2)
```

Two dataframes can be merged based on a shared column using the `merge()` function.

```{r}
dfr1 <- data.frame(x = 1:4, p = c("a", "b", "c","d"))
dfr2 <- data.frame(x = 3:6, q = c("l", "m", "n","o"))
dfr1
dfr2

merge(dfr1,dfr2,by="x")
merge(dfr1,dfr2,by="x",all.x=T)
merge(dfr1,dfr2,by="x",all.y=T)
merge(dfr1,dfr2,by="x",all=T)
```

# Data formats (Wide vs Long)

# Getting help

- Use `?function` to get function documentation
- Use `??bla` to search for a function
- Use `args(function)` to get the arguments to a function
- Go to the package CRAN page/webpage for vignettes
- [R Cookbook](http://www.cookbook-r.com/): General purpose reference.
- [Quick R](https://www.statmethods.net/): General purpose reference.
- [Stackoverflow](https://stackoverflow.com/): Online community to find solutions to your problems.

# Session info

```{r,echo=FALSE}
sessionInfo()
```

__End of document__
