---
title: "<i class='fas fa-pencil-alt' style='margin-left:4px;margin-right:4px;'></i> Precourse"
output:
  bookdown::html_document2:
    toc: false
    number_sections: false
    theme: flatly
    highlight: tango
    df_print: default
    code_folding: "none"
    self_contained: false
    keep_md: false
    encoding: 'UTF-8'
    css: "assets/lab.css"
---

```{r,child="assets/header-lab.Rmd"}
```

These are absolutely necessary for the course, if you do not already have a working R environment with R studio:

<div class="instruction">
<i class="fas fa-exclamation-circle"></i> Download and install the latest version of R from [r-project.org](https://cloud.r-project.org/). Install Rtools from [https://cran.r-project.org/bin/](https://cran.r-project.org/bin/). Rtools may be needed for installing new packages from source. 

<i class="fas fa-exclamation-circle"></i> Install [RStudio](https://www.rstudio.com/products/rstudio/download/). RStudio provides you with tools like code editor with highlighting, project management, version control, package building, debugger, profiler and more.
</div>

The following are the steps that are recommended for the course participants.

* As a pre-requisite to this course, all the participants have been asked to take the course: [Introduction to R](https://www.biology.lu.se/education/phd-studies/geneco-graduate-research-school-in-genomic-ecology/courses-workshops). 

* Although you would have a small session on how to handle data in R, it is advisible for you to freshen up your memory on data-types such as `data.frame`, `matrix` and `read/write files` and so on. 

* Some warm-up exercises in R if you are interested can be found  [here](https://immuntech.github.io/Bioinformatics_Workshop_ImmTech/Data/2019-10-15/R_exercise_Lokesh.html)



* The following R packages will be used in the course exercises. It might save you some time if you have these installed in advance.

```{r,eval=FALSE,block.title=TRUE}
# install R packages from CRAN
install.packages(c("ggplot2","ggtree","tidyverse","ggmap","pheatmap","devtools","cowplot",
                   "reshape2", "dplyr", "wesanderson", "scales", "ggthemes", "ggrepel"),dep=TRUE)
```

**End of document.**
